How to write a PhD in Biological Sciences: a guide for the uninitiated

### 2021-07-20 2.08-citations.Rmd
Add files via upload
# After copy edits

> Isaac Newton (in a letter to Robert Hooke in 1675) wrote:      
> **"If I have seen further it is by standing on the shoulders of Giants."**

Although the phrase is often attributed to Isaac Newton, it turns out that it was well accepted in Newton's time and has been traced back to the 12th century.

Essentially, standing on the shoulders of giants (Figures \@ref(fig:Googlescholar) is a recognition that all research is built upon research that has gone before it, and this is the basis for citations in the text of scientific papers. Patrick Dunleavy [@dunleavy2017citations] argues that citations are required to meet 7 criteria with respect to academic writing. Others have suggested that there are more criteria [see @harwood2009interview], and that these are a fundamental premise of academic work. But it was not always thus. If we dwell a moment longer, we can look back at the World's longest running scientific journal: *Philosophical Transactions of the Royal Society.*

You can download the first of those papers of 1665 and note that there are no citations (other than to books or letters) because there were no previously published articles from which to draw. However, even then authors noted that ideas came from previous authors and we can regularly find acknowledgements that refer back to Aristotle (~350 BC).

@@

Citations need to be used to back up any statements of fact that you make in your PhD chapter. Any examples need citations for where the examples came from, and when you make [arguments](argument.html), you add credibility to your arguments from each side by citing that these are from published works.

@@

I've seen many students try to write their text first and then only later fill in citations. I've had students hand in thesis chapters and manuscripts that have [ref] written strategically at the end of many sentences. I've even had the experience of doing this myself when I was a early career researcher (I probably still have those manuscripts somewhere because it's a sure way of never getting anything finished). I suggest that you **do not do this.** The reason is that if you write first and try to insert citations later, you are going to spend lots of time looking for the citation that fits your text. This is clearly not the right way around. Your text should be based on what's present in the literature.

I was planning to write that no-one would ever write an entire manuscript without references, and then put them all in at the end. Then I came across a video by [Pat Schloss](https://youtu.be/yk54_j3-4RM), and realised that I'm wrong [@schloss2018riffomonas]. Pat states early on (in the video) that he prefers to write first and cite later. He's clearly a very experienced researcher and knows his literature very well. Indeed, he knows it all so well that he already knows what citations fit what statements without needing to look them up. However, later on in the video Pat admits that even he sometimes gets it wrong, and that means that he ends up having to rewrite text when he finds the publication that he was thinking of and notes that what they actually found did not fit what he wrote. Although I can identify with what Pat is saying, I would find this re-writing far too time consuming to take this approach myself. I am familiar with most of the pertinent literature in my specialist area, but I'd rather check things before I start writing than leave it until later. I find that my memory is not only wrong about what other people have found, but also about what's in my own work. Therefore, I'd suggest that while you may aspire to write like Pat later on in your career, right now it'd be better for you to plan your writing with citations in mind, and cite while you write (not afterwards).

@@

## Citation Styles

@@

### Vancouver Style

Vancouver style dates back to medical journals in the 1970s and refer to conventions to avoid fraud in medical sciences. You are probably familiar with it even if you don't know the name. It is the convention that uses small superscript numbers to denote the place of citations in the main text, and the full reference given at the end of the document is provided in order of the citations (and their numbers), instead of alphabetical order. In biological sciences, Vancouver style is not used as widely as Harvard style, but is chosen as the superscript numbers tend not to impede the flow of the text for the reader.

Each style has its pros and cons. Vancouver style is more equitable and anonymous, as all cited people are represented only in the references and do not have their names plastered throughout the text. The diminutive numbers take up less space, allowing the reader and writer to concentrate more on their prose. On the other hand, if you are familiar with the literature, or you are interested in learning more about the literature, you’ll find that the little numbers take much more time to cross reference with the list at the end, than does Harvard style. Reference strings in Vancouver style can either be separate numbers, or numbers listed as a series. Because citations are numbered in the order in which they first mentioned, citation strings later in documents often end up as a list of non-sequential numbers separated by commas.

@@

The corresponding references in [Vancouver](vancouver.html) format are in the section on [references](references.html).

### Harvard Style

The standard way is to use the name and year in parentheses at the end of the statement to which the citation is relevant. Names of three or more authors are frequently reduced to “et al.” (often written in italics) which is short for the Latin et alia, meaning “and others”. References strings in Harvard style can go on for an entire line of text or more. Some journals have rules that mean on first mention the citation should have all authors (up to a certain number). This can become tedious when citations start taking up more space than text.

@@

You can find the corresponding references in [Harvard](harvard.html) format in the [references](references.html) section.

@@

Moving from Harvard to Vancouver style means that every time you would have had the year and date, you will replace this with a superscript denoting the number for that particular citation. However, Vancouver style does not allow for much variation in how you cite. You can’t use (e.g. ^1,2^) or (see ^3^ for a review). This means that when you originally wrote for a Harvard style journal, but then change the manuscript to submit it to a journal that uses Vancouver style, you’ll need to remove any explanations of citations. That doesn’t mean removing the citations themselves, but it may mean re-writing some sentences. Simply changing your reference manager to use the different referencing style will only be the start of your work.

@@

The impact of all invasive amphibians is similar to that of invasive birds and mammals [@measey2016global].

You’ll see that sometimes the names are brought into the sentence and become the central agents of the text. The phrase et alii (often abbreviated to et al.), literally means "and others" referring to the other authors (typically used when there are more than two - but check the journal format!).

Sometimes, instead of "et al." you can write "and colleagues" or "and others". This is something to do occasionally when you are looking to diversify some text. Don't over do it though.

@@

A citation string is a list of two or more citations that all assert the statement given. When you provide a string of citations you will need to decide which one comes first. Normally, this is given in the order of precedence: the oldest citation comes first, the youngest last. However, journals have their own styles and these may dictate how citation strings are ordered. For example, they could be alphabetical. If your referencing software has a style then you can relax. Otherwise, you'll need to look it up!

@@

Taxonomists have special rules for this, and I'll do it in another chapter (see section on [Scientific names and taxonomic authorities](taxonomy.html)). These are not the same as regular citations (because they don't appear in the literature cited and you don't have to have read the descriptions), and only some journals ask for them.

@@

Yes. One of the most common ways in which students mis-cite a paper is to use statements made in the introduction (or discussion) which were not the subject of the study. For example, in the introduction of their paper, Measey et al. [-@measey2016global] make comments on amphibian decline.

>Amphibian populations are currently declining across the globe [@wake2008are; @collins2009extinction; @pimm2014biodiversity] and alien amphibians are at least partially driving these declines through competition [@kupferberg1997bullfrog], hybridization [@dufresnes2015introgressive] and introduction of novel pathogens [@berger1999chytrid; @daszak2003infectious; @lamarca2005catastrophic; @martel2013batrachochytrium].

@@

Another common mistake is to forget which paper has which information. You can try to make sure that you don’t do this by taking better notes or a more accurate plan. And do the citations before or as you write, not afterwards!

@@

This does depend on the journal you are writing for. Some journals don’t permit citations to unpublished data or web sites. My suggestion would be to avoid such sources anyway (with particular exceptions – see below), unless it is really important that you include it. Other examples of texts to avoid are: text books (use the original paper instead), newspapers or magazines, blog posts, facebook pages (or other social media sites), predatory journals (or any non-peer reviewed article). These are all examples of __grey literature__.

There are of course exceptions. When you are writing about social media sites, or newspaper coverage, you will want to cite those sources.

I don't like citations to guide-books or general text books. This has become very common, but really smacks of laziness. Most of what's written in a guidebook has already been documented in the scientific literature, but guide-books generally don't provide sources for the information that they provide. Thus, it's easy to look up and find something in a guide-book, but suggests (to me) that you haven't spent enough time or effort reviewing the literature. Some guide-books are excellent and the authors have gone to a lot of trouble to incorporate original data and observations. But this is unusual, and most data can be found in publications.

One exception to the not citing websites relevant to our work is the IUCN Red List. Note that all entries on this site now have DOIs, and this might be a good guide for what is available to cite. The Digital Object Identifier ([DOI](doi.html)) is very useful as it means that there is a consistent record of that version. Otherwise, you could cite any website (even this one) and then the owner can go and change the site and it no longer says what you thought it did. The DOI removes this problem; there will always be an archived version with that particular DOI.

@@

The primary literature are studies or experiments that are done in order to test a hypothesis. Secondary literature includes reviews, syntheses or meta-analyses. Primacy (see below) is important, but this depends on the space you have and whether the review contains all the information you need to cite. Sometimes reviews (especially meta-analyses) provide more information. It’s preferable to use primary literature, but sometimes reviews (or meta-analyses) are actually more expedient to use, especially if they are not the focus of your study. You can even cite both when relevant.

## What is primacy and why does it matter?

Primacy in the literature is the paper/s where the authors provide the original idea or concept that you are citing. It is important in that you should give credit to original ideas over those who copy or simply repeat them, or even those who review them. Sometimes primacy is less important, especially if the concept is well known and has changed substantially. Then you should cite the most recent work that shows the ideas that you are wanting to show. Other times, primacy is important - especially when few have built on ideas or concepts since they were put forward. If in doubt place a citation to the original study and the most recent study, then flag this using a comment for your advisor to consider. The important point here is for you to give credit where it's due and not to overlook those who put in the hard work to publish original content.

## How many citations is enough?

@@

If you are looking for multiple examples of a statement that you’ve made and there are many possible, I’d suggest that you aim to produce three. Make sure that you use a suffix (like “e.g.”) to show that you are aware that these are examples of a widely reported phenomenon. You can choose these as you like, but may want to consider using what you consider to be ‘the best’ examples, and/or references that you are planning to use elsewhere in your paper/chapter. This can drive the total number down considerably, and helps to keep the citations more relevant to your work.

@@

If you are publishing relevant and appropriate papers, then there is no reason not to cite yourself. In many cases (such as with your thesis work), your own publications are likely to be more relevant to methodology and subject matter than much of the other work that is out there. However, if you’ve previously published on termite fungus and now you’re publishing on frog toes, it’s unlikely to be relevant. Choosing citations should include your comfort with being transparent. If you know when citing yourself that citing another article would be more correct in showing where an idea came from, but citing your own is more relevant to the application, then you should feel comfortable in citing both.

@@

Take a look at this article on self-citation. They claim that self-citations in the Natural Sciences run at 33% [@vanraan2007scoping], which provides you with a nice idea of what is acceptable. Remember that this would include not only your citations to your articles, but all citations to your co-authors articles too.

@@

In these days of scientometrics, we do need to acknowledge that citations act as a kind of currency. They count towards your H-index [@hirsch2005index] and this can reflect on your prospects as a postdoc or employee. What’s also clear is that cited papers get cited more, so it could really help your friends if you cite them. Obviously, the inverse is also true, so beware of the politics of citing. However, the most important points have already been raised, but will be tackled [in detail later](impactfactor.html). The study must be relevant and appropriate before it gets included as a citation in your work.

@@

Papers in journals with high impact factors are more likely to be cited because their contents are already thought to be of interest to a wide range of people. Indeed, there is evidence that identical statements published in many journals, those with higher Impact Factors are cited more [@perneger2010citation]. That is something that the editors of that journal will have considered even before sending the manuscript to review. Sometimes (but not always) the impact factor of the journal can be an indication of the quality of the study. But you should judge this for yourself when you critically read the paper.

>>>>

> Isaac Newton (in a letter to Robert Hooke in 1675) wrote: 
>
> **"If I have seen further it is by standing on the shoulders of Giants."**

Although the phrase is often attributed to Isaac Newton, it turns out that it was a well accepted phrase in Newton's time and has been traced back to the 12th century.

Essentially, standing on the shoulders of giants (Figure \@ref(fig:Googlescholar) is a recognition that all research is built upon research that has gone before it, and this is the basis for citations in the text of scientific papers. Patrick Dunleavy [-@dunleavy2017citations] argues that citations are required to meet 7 criteria with respect to academic writing. Others have suggested that there are more criteria [see @harwood2009interview], and that these are a fundamental premise of academic work. But it was not always thus. If we dwell a moment longer, we can look back at the World's longest-running scientific journal: *Philosophical Transactions of the Royal Society.*

You can download the first of those papers of 1665 and note that there are no citations (other than to books or letters) because there were no previously published articles from which to draw (Figure \@ref(fig:Boyle)). However, even then, authors noted that ideas came from previous authors and we can regularly find acknowledgements that refer back to Aristotle (~350 BC).

@@

Citations need to be used to back up any statements of fact that you make in your PhD chapter. Any examples need citations for where the examples came from, and when you make [arguments](#argument), you add credibility to your arguments from each side by citing that these are from published works.

@@

I've seen many students try to write their text first and then only later fill in citations. I've had students hand in thesis chapters and manuscripts that have '[ref]' written strategically at the end of many sentences. I've even had the experience of doing this myself when I was an early career researcher (I probably still have those manuscripts somewhere because it's a sure way of never getting anything finished). I suggest that you **do not do this.** The reason is that if you write first and try to insert citations later, you are going to spend lots of time looking for the citation that fits your text. This is clearly not the right way around. Your text should be based on what's present in the literature.

I was planning to write that no-one would ever write an entire manuscript without references, and then put them all in at the end. Then I came across a video by [Pat Schloss](https://youtu.be/yk54_j3-4RM), and realised that I'm wrong [@schloss2018riffomonas]. Pat states early on (in the video) that he prefers to write first and cite later. He's clearly a very experienced researcher and knows his literature very well. Indeed, he knows it all so well that he already knows what citations fit what statements without needing to look them up. However, later on in the video, Pat admits that even he sometimes gets it wrong, and that means that he ends up having to rewrite text when he finds the publication that he was thinking of and notes that what they actually found did not fit what he wrote. Although I can identify with what Pat is saying, I would find this re-writing far too time consuming to take this approach myself. I am familiar with most of the pertinent literature in my specialist area, but I'd rather check things before I start writing than leave it until later. I find that my memory is not only wrong about what other people have found but also about what's in my own work. Therefore, I'd suggest that while you may aspire to write like Pat later on in your career, right now it'd be better for you to plan your writing with citations in mind, and cite while you write (not afterwards).

@@

## Citation styles

@@

### Vancouver style

Vancouver style dates back to medical journals in the 1970s and refers to conventions to avoid fraud in medical sciences. You are probably familiar with it, even if you don't know the name. It is the convention that uses small superscript numbers to denote the place of citations in the main text, and the full reference given at the end of the document is provided in order of the citations (and their numbers), instead of alphabetical order. In biological sciences, Vancouver style is not used as widely as Harvard style, but is chosen as the superscript numbers tend not to impede the flow of the text for the reader.

Each style has its pros and cons. Vancouver style is more equitable and anonymous, as all cited people are represented only in the references and do not have their names plastered throughout the text. The diminutive numbers take up less space, allowing the reader and writer to concentrate more on their prose. On the other hand, if you are familiar with the literature, or you are interested in learning more about the literature, you’ll find that the little numbers take much more time to cross-reference with the list at the end, than does Harvard style. Reference strings in Vancouver style can either be separate numbers, or numbers listed as a series. Because citations are numbered in the order in which they first mentioned, citation strings later in documents often end up as a list of non-sequential numbers separated by commas.

@@

The corresponding references in [Vancouver](#vancouver) format are in the section on [references](#references).

### Harvard Style {#Hstyle}

The standard way is to use the name and year in parentheses at the end of the statement to which the citation is relevant. Names of three or more authors are frequently reduced to 'et al.' (often written in italics: _et al._) which is short for the Latin _et alia_, meaning 'and others'. References strings in Harvard style can go on for an entire line of text or more. Some journals have rules that mean on the first mention, the citation should have all authors (up to a certain number). This can become tedious when citations start taking up more space than text.

@@

You can find the corresponding references in [Harvard](#harvard) format in the [references](#references) section.

@@

Moving from Harvard to Vancouver style means that every time you would have had the year and date, you will replace this with a superscript denoting the number for that particular citation. However, Vancouver style does not allow for much variation in how you cite. You can’t use (e.g. ^1,2^) or (see ^3^ for a review). This means that when you originally wrote for a Harvard style journal, but then change the manuscript to submit it to a journal that uses Vancouver style, you’ll need to remove any explanations of citations. That doesn’t mean removing the citations themselves, but it may mean re-writing some sentences. Simply changing your reference manager to use a different referencing style will only be the start of your work.

@@

>The impact of all invasive amphibians is similar to that of invasive birds and mammals [@measey2016global].

You’ll see that sometimes the names are brought into the sentence and become the central agents of the text.

Sometimes, instead of 'et al.' you can write 'and colleagues' or 'and others'. This is something to do occasionally when you are looking to diversify some text. Don't over do it though.

@@

A citation string is a list of two or more citations that all assert the statement given. When you provide a string of citations, you will need to decide which one comes first. Normally, this is given in the order of precedence: the oldest citation comes first, the youngest last. However, journals have their own styles and these may dictate how citation strings are ordered. For example, they could be alphabetical. If your referencing software has a style then you can relax. Otherwise, you'll need to look it up!

@@

Taxonomists have special rules for this, and this will be explained in another chapter (see section on [Scientific names and taxonomic authorities](#taxonomy)). These are not the same as regular citations (because they don't appear in the literature cited and you don't have to have read the descriptions), and only some journals ask for them.

@@

Yes. One of the most common ways in which students mis-cite a paper is to use statements made in the introduction (or discussion) which were not the subject of the study. For example, in the introduction of their paper, Measey et al. [-@measey2016global] make comments on amphibian decline (see [above](#Hstyle)).

@@

Another common mistake is to forget which paper has which information. You can try to make sure that you don’t do this by taking better notes or write a more accurate plan. And do the citations before or as you write, not afterwards!

@@

This does depend on the journal you are writing for. Some journals don’t permit citations to unpublished data or web sites. My suggestion would be to avoid such sources anyway (with particular exceptions – see below), unless it is really important that you include it. Other examples of texts to avoid are: text books (use the original paper instead), newspapers or magazines, blog posts, Facebook pages (or other social media sites), predatory journals (or any non-peer-reviewed article). These are all examples of __grey literature__.

There are, of course, exceptions. When you are writing about social media sites, or newspaper coverage, you will want to cite those sources.

I don't like citations to guide-books or general text books. This has become very common, but really smacks of laziness. Most of what's written in a guidebook has already been documented in the scientific literature, but guide-books generally don't provide sources for the information that they provide. Thus, it's easy to look up and find something in a guide-book but suggests (to me) that you haven't spent enough time or effort reading the literature. Some guide-books are excellent and the authors have gone to a lot of trouble to incorporate original data and observations. But this is unusual, and most data can be found in publications.

One exception to not citing websites is the IUCN Red List. Note that all entries on this site now have DOIs, and this might be a good guide for what is available to cite. The Digital Object Identifier ([DOI](#DOI)) is very useful as it means that there is a consistent record of that version. Otherwise, you could cite any website and then the owner can go and change the site and it no longer says what you thought it did. The DOI removes this problem; there will always be an archived version with that particular DOI.

@@

The primary literature consists of studies or experiments that are done in order to test a hypothesis. Secondary literature includes reviews, syntheses or meta-analyses. Primacy (see below) is important, but this depends on the space you have and whether the review contains all the information you need to cite. Sometimes reviews (especially meta-analyses) provide more information. It’s preferable to use primary literature, but sometimes reviews (or meta-analyses) are actually more expedient to use, especially if they are not the focus of your study. You can even cite both when relevant.

## What is rule of priority and why does it matter?

The rule of priority (or primacy) in the literature is that those paper/s where the authors first provide the original idea or concept, or evidence for this are the ones that are cited [@strevens2003role]. Papers that are published afterwards, are simply seen as repeating the same story as those who published first, even if the time-lines of the research itself don't reflect this. The rule of priority makes sense, in that we have an obligation to name the shoulders on which we stand (Figures \@ref(fig:Googlescholar). Thus, it is important in that you should give credit to original ideas over those who copy or simply repeat them, or even those who review them. Tracing back the origins of ideas makes for interesting reading and a deeper understanding of your subject.

Sometimes primacy is less important, especially if the concept is well known and/or has changed substantially. Then you should cite the most recent work that shows the ideas that you are wanting to show. Other times, primacy is more important - especially when few have built on ideas or concepts since they were put forward, or in the [taxonomic literature](taxonomy.html). If in doubt, place a citation to the original study and the most recent study, then flag this using a comment for your advisor to consider. The important point here is for you to give credit where it's due and not to overlook those who put in the hard work to publish original content.

Some have argued that the rule of priority is of benefit to science as it drives competition and this spurs creativity. On the other hand, the winner-takes-all approach of being first with something drives secrecy and dishonesty, and does not genuinely reflect the team effort that is the [science project](lifescientific.html) [@casadevall2012reforming].

## How many citations are enough?

@@

If you are looking for multiple examples of citations for a statement that you’ve made and there are many possibilities, I’d suggest that you aim to produce three. Make sure that you use a suffix (like 'e.g.') to show that you are aware that these are examples of a widely reported phenomenon. You can choose these as you like, but may want to consider using what you consider to be ‘the best’ examples, and/or references that you are planning to use elsewhere in your paper/chapter. This can drive the total number of citations down considerably, and helps to keep the citations more relevant to your work.

@@

If you are publishing relevant and appropriate papers, then there is no reason not to cite yourself. In many cases (such as with your thesis work), your own publications are likely to be more relevant to methodology and subject matter than much of the other work that is out there. However, if you’ve previously published on termite fungus and now you’re publishing on rat toes, it’s unlikely to be relevant. Choosing citations should include your comfort with being transparent. If you know when citing yourself that citing another article would be more correct in showing where an idea came from, but citing your own is more relevant to the application, then you should feel comfortable in citing both.

@@

Take a look at this article on self-citation. They claim that self-citations in the Natural Sciences run at 33% [@vanraan2007scoping], which provides you with a nice idea of what is acceptable. Remember that this would include not only your citations to your articles, but all citations to your co-authors' articles too.

@@

In these days of scientometrics, we do need to acknowledge that citations act as a kind of currency. They count towards your H-index [@hirsch2005index] and this can reflect on your prospects as a postdoc or employee. What’s also clear is that cited papers get cited more, so it could really help your friends if you cite them. Obviously, the inverse is also true, so beware of the politics of citing. However, the most important points have already been raised. The study must be relevant and appropriate before it gets included as a citation in your work.

@@

Papers in journals with high impact factors are more likely to be cited because their contents are already thought to be of interest to a wide range of people. Indeed, there is evidence that for identical statements published in many journals, those with higher Impact Factors are cited more [@perneger2010citation]. Sometimes (but not always) the impact factor of the journal can be an indication of the quality of the study. But you should judge this for yourself when you critically read the paper.
